Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys

### 2019-03-08 doc/article.tex
clean up to fit page limit

%\documentclass{aastex62}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb,amsmath}
\usepackage{natbib,verbatim}
\usepackage[multiple]{footmisc}
\usepackage{aas_macros}
%\usepackage{grffile}
%\usepackage{textcase}
\usepackage{hyperref}
%\usepackage{geometry}

%\usepackage[overload]{textcase}

@@

Simone Ferraro\footnotemark[1],

@@

S.~Perlmutter\footnotemark[1]\textsuperscript{,}\footnotemark[20],
\footnotetext[20]{Department of Physics, University of California Berkeley, Berkeley, CA 94720}

@@

A.~Shafieloo\footnotemark[17],
\footnotetext[17]{Korea Astronomy and Space Science Institute, Yuseong-gu, Daedeok-daero 776, Daejeon 34055, Korea}
A.~Slo\u{z}ar\footnotemark[18],
\footnotetext[18]{Brookhaven National Laboratory, Physics Department, Upton, NY

@@

@@

M.~Wood-Vasey\footnotemark[19]
\footnotetext[19]{PITT PACC, Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA 15260, USA} 

@@

\textbf{Endorsers:}
 Behzad Ansarinejad $^{1,2}$, 
 Robert Armstrong $^{}$, 
 Carlo Baccigalupi $^{3,4,5}$, 
 Maciej Bilicki $^{6}$, 
 Julian Borrill $^{7}$, 
 Simone Ferraro $^{7}$, 
 Llu\'is Galbany $^{8}$, 
 Juan Garc\'ia-Bellido $^{9}$, 
 Juan Garc\'ia-Bellido $^{10}$, 
 Martina Gerbino $^{}$, 
 Mandeep S.S. Gill $^{11,12,13}$, 
 Larry Gladney $^{14}$, 
 Marc Kamionkowski $^{1,19}$, 
 Kazuya Koyama $^{20}$, 
 Michele Liguori $^{21}$, 
 Joel Meyers $^{22}$, 
 "Surhud More" $^{23}$, 
 Jeffrey A. Newman $^{8}$, 
 Gustavo Niz $^{24}$, 
 "Antonella Palmese" $^{25}$, 
 Francesco Piacentini $^{26}$, 
 Francesco Piacentni  $^{26,27}$, 
 Steven Rodney $^{28}$, 
 Benjamin Rose $^{29}$, 
 Lado Samushia $^{30}$, 
 Neelima Sehgal $^{31}$, 
 Leonardo Senatore $^{11}$, 
 Huanyuan Shan $^{1,33}$, 
 Scott Watson $^{35}$, 
 Martin White $^{36,7}$, 
 Weishuang Xu $^{37}$, 
 Gong-Bo Zhao $^{38,39,20}$
  \linebreak

@@

@@

%\begin{abstract}
%In the upcoming decade cadenced wide-field imaging surveys 
%will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
%hundreds of thousands.  The increase in the number density  and solid-angle coverage 
%of SNe~Ia, in parallel with improvements in the standardization of
%their absolute magnitudes, now make them competitive probes of the growth of structure and hence of gravity.  The peculiar velocity power spectrum
%is sensitive to $\gamma$, which captures the effect of gravity on the linear growth of structure through the relation $f=\Omega_M^\gamma$.
%In the next decade the peculiar velocities of
%SNe~Ia in the local $z<0.3$ Universe will provide a measure of $\gamma$ with as low as
%$0.01$ precision that can definitively distinguish  between General Relativity and leading models of alternative gravity.
%\end{abstract}
%\section{Introduction}

{\bf Introduction:} In the late 1990's, Type~Ia supernovae (SNe~Ia) were used as distance probes to measure the homogeneous expansion history of the Universe.  The remarkable discovery
that the expansion is accelerating  has called into question our basic understanding of the gravitational forces within the Universe.  Either it
is dominated by a ``dark energy'' that is gravitationally repulsive, or General Relativity is inadequate and needs to be replaced by a modified theory of
gravity.  It is only appropriate that in the upcoming decade, with their sheer numbers, solid-angle coverage,
and improved distance precisions, SNe~Ia will provide measurements of the {\it inhomogeneous} motions of structures in the Universe
that will provide an unmatched test of whether dark energy or modified gravity is responsible for the accelerating expansion of the Universe.

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure  the influence of gravity on structure formation within the Universe.
The SN~Ia Hubble diagram has  scatter along the redshift axis since peculiar velocities induce differing observed and cosmological redshifts.  The contribution
of this scatter is pronounced at low redshifts and when the scatter along the magnitude axis (i.e.\ intrinsic magnitude dispersion) is small.
The peculiar velocity power spectrum is sensitive to the growth of structure as $P_{vv}\propto (fD)^2$, where $D$ is  the spatially-independent
growth factor for the linear evolution of density perturbations and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate with $a$ the scale factor  \cite{2006PhRvD..73l3526H,2011ApJ...741...67D}.
%\footnote{
%To be precise, the peculiar
%velocity power spectrum also depends on the Hubble parameter as $P_{vv}\propto (HfD)^2$.  A supernova  survey  measures luminosity distance fluctuations
%$\delta_{d_L} = (d_L- \bar{d}_L(z))/\bar{d}_L(z)$, where $d_L$ and  $\bar{d}_L(z)$ are the measured and expected distances at the observed redshift $z$.
%To first order in peculiar velocity along the line of sight $v$, $\delta_{d_L} =  v \left(1-\frac{1}{H\bar{d}_L(z)}\right) \approx -\frac{v}{H\bar{d}_L(z)}$
%at low redshift.   The $H$-dependences of  $P_{vv}$ and the conversion from distances to velocities cancel, making  peculiar velocity surveys sensitive to $(fD)^2$.
%}
The $\Lambda$CDM prediction for the $z=0$ peculiar velocity power spectrum is shown in Figure~\ref{power:fig}. The growth of structure depends on gravity, indeed 
\cite{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravity follow the relation
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively \cite[see][for a review of
these gravity models within a cosmological context]{HUTERER201523}.  Adopting this parameterization to model gravity, peculiar velocity
surveys are  sensitive to $fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$, whose prediction for $\Lambda$CDM  is plotted 
in Figure~2 of  \cite{1475-7516-2013-04-031}. (Note that we prefer the notation $fD$ in place of $f\sigma_8$, which is commonly used in the literature.)
%
%The parameter $\sigma_8$, the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres, is 
%commonly used in place of $D$ to normalize the
%overall amplitude of  overdensities, so the standard parameterization used by the community is $f\sigma_8$.}

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted peculiar velocity power spectrum $k^3P_{vv}(z=0)$ for $\mu=\cos{(\hat{k} \cdot \hat{r})}=1, 0.5$ (magenta, cyan) 
($\hat{r}$ is the line of sight), as predicted for General Relativity in the linear regime.
Overplotted are volume-weighted peculiar-velocity power-spectrum shot noise  (diagonal lines) for various observing parameters.  Red shows the shot noise expected from a 2-year LSST survey
while black shows a 10-year LSST survey.  The dotted and dashed lines indicate the assumed intrinsic magnitude dispersion, using 0.08 (dashed) or 0.15 mag (dotted).  The expected shot
noise from TAIPAN is shown in green (dash-dotted). 
%
%Overplotted are volume-weighted peculiar velocity shot-noise  $k^3\sigma^2/n$ at $z=0.1$ expected from 2- and 10-year (red, black) supernova densities, 0.08 and 0.15~mag (dashed, dotted) intrinsic magnitude dispersions, and TAIPAN (green, dash-dotted).
The bottom solid grey horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{\rm max}$.
\label{power:fig}}
\end{figure}

Peculiar velocity surveys have already been  used to measure $fD$, though not to a level where  gravity models can be precisely distinguished.
 \cite{2017MNRAS.471..839A} use 6dFGS peculiar velocities using the Fundamental Plane distances of elliptical galaxies, to estimate absolute magnitudes
 with
 $\sim 0.43$~mag  precision to
get a 15\% uncertainty in $fD$ at $z\approx 0$ using combined density and velocity information.
The upcoming 
TAIPAN survey \cite{2017PASA...34...47D} will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys \cite{2008ExA....22..151J} will obtain Tully-Fisher distances (based on the $\sim 0.48$~mag calibration of absolute magnitude based on the  HI 21cm line width)
of galaxies with densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$ covering 75\% of the sky.
These surveys combined are projected to have 3\% $fD$ uncertainties \cite{2017MNRAS.464.2517H}.
For reference, DESI projects 10\% precision of $fD$ at $z \approx 0.3$  by statistically looking 
for expected signatures of galaxies infalling toward mass overdensities, Redshift Space Distortions (RSD).
Relative to galaxies with  Fundamental Plane or Tully-Fisher distances, 
SN~Ia host galaxies currently have significantly lower number density but have better per-object peculiar velocity precision.
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\cite{PhysRevLett.99.081301,2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
SNe~Ia discovered by ZTF over the next several years will provide first probative measures of $fD$ at $z<0.1$.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \cite{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter spectral model yields  absolute magnitude
dispersion $\sigma_M \gtrsim 0.12$~mag.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SN absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \cite[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}. 
Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.  If corrections of all SNe are not possible, the use of SN~Ia subclasses is an option though at the expense of reducing the sample.
At this precision the intrinsic velocity dispersion  at $z=0.028$ is  of $300$~km\,s$^{-1}$, i.e.\ a single SN~Ia  is of such quality as to
provide a peculiar velocity with $S/N \sim 1$.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands; over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$.
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.  With similar densities,
 the (two) 10-year SN~Ia survey will have
 a (6) 29$\times$ reduction in shot-noise, $\sigma^2_M/n$, relative to the Fundamental Plane survey of TAIPAN.

Given these two advances, supernovae discovered by wide-field searches in the next decade will 
 be able to stringently constrain the growth of structure in the low-redshift Universe.
For example, 
over the course of a decade a SN survey relying on LSST discoveries 
plus spectroscopic redshifts can produce  4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, cumulatively giving
2.2\% uncertainty on $fD$  within this interval, where at
$0<z<0.2$ most of the probative power comes from peculiar velocities and at higher redshifts from Redshift Space Distortions  \cite{2017ApJ...847..128H}.

%\section{Testing Gravity with Peculiar Velocity Surveys}
{\bf Testing Gravity with Peculiar Velocity Surveys:} While the growth rate $fD$ can be used to test several aspects of physics beyond the standard cosmological model (e.g.\ dark matter clustering, dark energy evolution), our scientific interest is in probing gravity,
so here we focus  on the growth index $\gamma$.
To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega_M + \int \Omega_M^{\gamma} \ln{\Omega_M}\, d\ln{a} \approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively
 in $\Lambda$CDM;  two surveys with the same fractional precision  in $fD$ will have different precision in $\gamma$,
 with the one at lower redshift providing the tighter constraint. 
 In this section, we focus on how well peculiar velocity surveys in the upcoming decade can measure $\gamma$ and the sensitivity of the measurement
 on survey parameter choices.

We project uncertainties  on the growth index, $\sigma_\gamma$ for a suite of idealized surveys
using a Fisher matrix analysis similar to that of \cite{2017ApJ...847..128H, 2017MNRAS.464.2517H}
\cite[there is an alternative approach using an estimator for the mean pairwise velocity][]{2011PhRvD..83d3004B}.
The Fisher information matrix is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_{r_{\rm min}}^{r_{\rm max}}  \int_{k_{\rm min}}^{k_{\rm max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}
\label{cov:eq}
\end{equation}
and the parameters considered are $\lambda \in \{\gamma,bD, \Omega_{M_0}\}$.
The parameter dependence enters through $(fD)(\gamma)$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy count overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.  While the $bD$ term does contain information on $\gamma$, its constraining power is not used here.
Both $f$ and $D$ depend on $\Omega_M=\frac{\Omega_{M_0}}{\Omega_{M_0} + (1-\Omega_{M_0})a^3}$.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered here.

The uncertainty  $\sigma_\gamma$ of a survey depends on its solid angle $\Omega$, depth
given by the comoving distance out to the maximum redshift $r_{\rm max}=r(z_{\rm max})$, duration
$t$ through $n=\epsilon \phi t$ where $\phi$ is the observer-frame SN~Ia rate
and $\epsilon$ is the sample-selection efficiency, and
the intrinsic SN~Ia magnitude dispersion 
through the resulting peculiar velocity intrinsic dispersion $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$.
An analysis in which the RSD and velocity surveys are treated independently 
sets the off-diagonal elements of $C$ to zero.

We consider SN peculiar velocity surveys for a range of redshift depths $z_{\rm max}$ for survey durations of $t=2$ and 10~years.
The other survey parameters
$\Omega=3\pi$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
The $k$-limits are  taken to be $k_{\rm min} = \pi/r_{\rm max}$  and 
$k_{\rm max} = 0.1$~$h$\,Mpc$^{-1}$.
A minimum distance $r_{\rm min}=r(z=0.01)$ is imposed as the physics derivation is taken at the limit
where peculiar velocities are significantly  smaller than
the cosmological redshift.
The sample-selection efficiency $\epsilon$ is redshift-independent, i.e.\ the native redshift distribution is not sculpted.
The input bias of SN~Ia host galaxies is set as $b=1.2$.  An independent measurement of $\Omega_{M0}=0.3\pm0.005$
is included and is important contributor to the $\gamma$ constraint.  Number densities are taken to be
direction-independent, neglecting the slight declination-dependence of SN-survey time windows. Code used for the analysis in this article available
at \url{http:tiny.cc/PVScience}.

>>>>

%\documentclass{aastex62}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb,amsmath}
\usepackage{natbib,verbatim}
\usepackage[multiple]{footmisc}
\usepackage{aas_macros}
%\usepackage{grffile}
%\usepackage{textcase}
\usepackage{hyperref}
%\usepackage{geometry}
\usepackage{titlesec}

\titleformat*{\section}{\large\bfseries}
%\usepackage[overload]{textcase}

@@

S.~Ferraro\footnotemark[1],

@@

S.~Perlmutter\footnotemark[1]\textsuperscript{,}\footnotemark[17],
\footnotetext[17]{Department of Physics, University of California Berkeley, Berkeley, CA 94720}

@@

A.~Shafieloo\footnotemark[18],
\footnotetext[18]{Korea Astronomy and Space Science Institute, Yuseong-gu, Daedeok-daero 776, Daejeon 34055, Korea}
A.~Slo\u{z}ar\footnotemark[19],
\footnotetext[19]{Brookhaven National Laboratory, Physics Department, Upton, NY

@@

R.~B.~Tully\footnotemark[20],
\footnotetext[20]{Institute for Astronomy, University of Hawaii, 2680 Woodlawn Drive, Honolulu, HI 96822, USA}

@@

M.~Wood-Vasey\footnotemark[21]
\footnotetext[21]{PITT PACC, Department of Physics and Astronomy, University of Pittsburgh, Pittsburgh, PA 15260, USA} 

@@

\textbf{Endorsers:}
 \linebreak

@@

We present the first  projections for the precision in $\gamma$ for a range of realistic SN peculiar-velocity survey scenarios.

@@

%\begin{abstract}
%In the upcoming decade cadenced wide-field imaging surveys 
%will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
%hundreds of thousands.  The increase in the number density  and solid-angle coverage 
%of SNe~Ia, in parallel with improvements in the standardization of
%their absolute magnitudes, now make them competitive probes of the growth of structure and hence of gravity.  The peculiar velocity power spectrum
%is sensitive to $\gamma$, which captures the effect of gravity on the linear growth of structure through the relation $f=\Omega_M^\gamma$.
%In the next decade the peculiar velocities of
%SNe~Ia in the local $z<0.3$ Universe will provide a measure of $\gamma$ with as low as
%$0.01$ precision that can definitively distinguish  between General Relativity and leading models of alternative gravity.
%\end{abstract}
\section{Introduction}
In the late 1990's, Type~Ia supernovae (SNe~Ia) were used as distance probes to measure the homogeneous expansion history of the Universe.  The remarkable discovery
that the expansion is accelerating  has called into question our basic understanding of the gravitational forces within the Universe.  Either it
is dominated by a ``dark energy'' that is gravitationally repulsive, or General Relativity is inadequate and needs to be replaced by a modified theory of
gravity.  It is only appropriate that in the upcoming decade, with their sheer numbers, solid-angle coverage,
and improved distance precisions, SNe~Ia will provide measurements of the {\it inhomogeneous} motions of structures in the Universe
that will provide an unmatched test of whether dark energy or modified gravity is responsible for the accelerating expansion of the Universe.

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure  the influence of gravity on structure formation within the Universe.
Peculiar velocities induce scatter along the redshift axis of the SN Hubble diagram, which is
pronounced at low redshifts and when the magnitude scatter (e.g.\ due to intrinsic magnitude dispersion) is small.
The peculiar velocity power spectrum is sensitive to the growth of structure as $P_{vv}\propto (fD)^2$, where $D$ is  the spatially-independent
``growth factor'' in the linear evolution of density perturbations and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate where $a$ is the scale factor  \cite{2006PhRvD..73l3526H,2011ApJ...741...67D}.
%\footnote{
%To be precise, the peculiar
%velocity power spectrum also depends on the Hubble parameter as $P_{vv}\propto (HfD)^2$.  A supernova  survey  measures luminosity distance fluctuations
%$\delta_{d_L} = (d_L- \bar{d}_L(z))/\bar{d}_L(z)$, where $d_L$ and  $\bar{d}_L(z)$ are the measured and expected distances at the observed redshift $z$.
%To first order in peculiar velocity along the line of sight $v$, $\delta_{d_L} =  v \left(1-\frac{1}{H\bar{d}_L(z)}\right) \approx -\frac{v}{H\bar{d}_L(z)}$
%at low redshift.   The $H$-dependences of  $P_{vv}$ and the conversion from distances to velocities cancel, making  peculiar velocity surveys sensitive to $(fD)^2$.
%}
The $\Lambda$CDM prediction for the $z=0$ peculiar velocity power spectrum is shown in Figure~\ref{power:fig}. The growth of structure depends on gravity;
\cite{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravity follow the relation
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively (see \cite{HUTERER201523} for a review
or these  models).  Using this parameterization to model gravity, peculiar velocity
surveys probe $\gamma$ through $fD$, whose $\gamma$-dependence is plotted 
in Figure~2 of  \cite{1475-7516-2013-04-031}.
%$fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$
%
%The parameter $\sigma_8$, the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres, is 
%commonly used in place of $D$ to normalize the
%overall amplitude of  overdensities, so the standard parameterization used by the community is $f\sigma_8$.}

\begin{figure}[h]
\centering
\includegraphics[width=0.49\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted peculiar velocity power spectrum $k^3P_{vv}(z=0)$ for $\mu \equiv \cos{(\hat{k} \cdot \hat{r})}=1, 0.5$ (magenta, cyan) 
where $\hat{r}$ is the line of sight, as predicted for General Relativity in the linear regime.
Overplotted are peculiar-velocity power-spectrum shot noise  (diagonal lines) for various observing parameters.  Red shows the shot noise expected from a 2-year LSST survey
while black shows a 10-year LSST survey.  The dotted and dashed lines indicate the assumed intrinsic magnitude dispersion, using 0.08 (dashed) or 0.15 mag (dotted).  The expected shot
noise from TAIPAN is shown in green (dash-dotted). 
%
%Overplotted are volume-weighted peculiar velocity shot-noise  $k^3\sigma^2/n$ at $z=0.1$ expected from 2- and 10-year (red, black) supernova densities, 0.08 and 0.15~mag (dashed, dotted) intrinsic magnitude dispersions, and TAIPAN (green, dash-dotted).
The bottom solid grey horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{\rm max}$.
\label{power:fig}}
\end{figure}

Peculiar velocity surveys have already been  used to measure $fD$ (also referred to as $f\sigma_8$), though not to a level where  gravity models can be precisely distinguished.
 \cite{2017MNRAS.471..839A} use 6dFGS peculiar velocities using  Fundamental Plane distances of elliptical galaxies to estimate absolute magnitudes
 with
 $\sim 0.43$~mag  precision, yielding a 15\% uncertainty in $fD$ at $z\approx 0$.
The upcoming 
TAIPAN survey \cite{2017PASA...34...47D} will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys \cite{2008ExA....22..151J} will obtain Tully-Fisher distances (based on the $\sim 0.48$~mag calibration of absolute magnitude based on the  HI 21cm line width)
of galaxies with densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$ covering 75\% of the sky.
These surveys combined are projected to have 3\% uncertainties in $fD$ \cite{2017MNRAS.464.2517H}.
For reference, DESI projects a 10\% precision of $fD$ at $z \approx 0.3$  by looking 
for signatures (Redshift Space Distortions; RSD) expected from galaxies infalling toward mass overdensities.
Relative to galaxies with  Fundamental Plane or Tully-Fisher distances, 
SN~Ia host galaxies currently have significantly lower number density but have better per-object peculiar velocity precision.
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\cite{PhysRevLett.99.081301,2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
SNe~Ia discovered by ZTF \cite{2019PASP..131a8002B} over the next several years will provide first probative measures of $fD$ at $z<0.1$.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \cite{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter spectral model yields  absolute magnitude
dispersion $\sigma_M \gtrsim 0.12$~mag.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SN absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \cite[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}. 
Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.  At this precision the intrinsic velocity dispersion  at $z=0.028$ is  of $300$~km\,s$^{-1}$, i.e.\ a single SN~Ia  is of such quality as to
measure a peculiar velocity with $S/N \sim 1$.
 If corrections of all SNe~Ia are not possible, the use of SN~Ia subclasses is an option though at the expense of reducing the
numbers of velocity probes.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae from the hundreds to the
hundreds of thousands; over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$.
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.  With similar densities,
 the (two) ten-year SN~Ia survey will have
 a (6) 29$\times$ reduction in shot-noise, $\sigma^2_M/n$, relative to the Fundamental Plane survey of TAIPAN.

Given these  advances, supernovae discovered by wide-field searches in the next decade will 
 be able to tightly constrain the growth of structure in the low-redshift Universe.
For example, 
over the course of a decade a SN survey relying on LSST discoveries 
plus spectroscopic redshifts can produce  4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, cumulatively giving
2.2\% uncertainty on $fD$  within this interval, where at
$0<z<0.2$ most of the probative power comes from peculiar velocities and at higher redshifts from RSD  \cite{2017ApJ...847..128H}.

\section{Testing Gravity with Peculiar Velocity Surveys}
While the growth rate $fD$ can be used to test several aspects of physics beyond the standard cosmological model (e.g.\ dark matter clustering, dark energy evolution), our scientific interest is in probing gravity,
so here we focus  on the growth index $\gamma$.
To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega_M + \int \Omega_M^{\gamma} \ln{\Omega_M}\, d\ln{a} \approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively
 in $\Lambda$CDM;  two surveys with the same fractional precision  in $fD$ will have different precision in $\gamma$,
 with the one at lower redshift providing the tighter constraint. 
 In this section, we demonstrate that peculiar velocity surveys in the upcoming decade can measure $\gamma$ precisely
for a range of survey-parameter choices.

We project uncertainties  on the growth index, $\sigma_\gamma$ for a suite of idealized surveys
using a Fisher matrix analysis similar to that of \cite{2017ApJ...847..128H, 2017MNRAS.464.2517H}
(there is an alternative approach using an estimator for the mean pairwise velocity
\cite{2011PhRvD..83d3004B}).
The ``cross-correlation'' analysis incorporates both galaxy overdensities and peculiar velocities.
The Fisher information matrix is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_{r_{\rm min}}^{r_{\rm max}}  \int_{k_{\rm min}}^{k_{\rm max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}
\label{cov:eq}
\end{equation}
and the parameters considered are $\lambda \in \{\gamma,bD, \Omega_{M_0}\}$.
The parameter dependence enters through $fD$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy count overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu\equiv \cos{(\hat{k} \cdot \hat{r})}$ where $\hat{r}$ is the direction of
the line of sight.  While the $bD$ term does contain information on $\gamma$, its constraining power is not used here.
Both $f$ and $D$ depend on $\Omega_M=\frac{\Omega_{M_0}}{\Omega_{M_0} + (1-\Omega_{M_0})a^3}$.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered here.

The uncertainty  $\sigma_\gamma$ of a survey depends on its solid angle $\Omega$, depth
given by the comoving distance out to the maximum redshift $r_{\rm max}=r(z_{\rm max})$, duration
$t$ through $n=\epsilon \phi t$ where $\phi$ is the observer-frame SN~Ia rate
and $\epsilon$ is the sample-selection efficiency, and
the intrinsic SN~Ia magnitude dispersion 
through the resulting peculiar velocity intrinsic dispersion $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$.
%An analysis in which the RSD and velocity surveys are treated independently 
%sets the off-diagonal elements of $C$ to zero.

We consider SN peculiar velocity surveys for a range of redshift depths $z_{\rm max}$ for durations of $t=2$ and 10~years.
The other survey parameters
$\Omega=3\pi$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
The $k$-limits are  taken to be $k_{\rm min} = \pi/r_{\rm max}$  and 
$k_{\rm max} = 0.1$~$h$\,Mpc$^{-1}$.
A minimum distance $r_{\rm min}=r(z=0.01)$ is imposed as our analysis assumes that peculiar velocities are significantly  smaller than
the cosmological redshift.
The sample-selection efficiency $\epsilon$ is redshift-independent, i.e.\ the native redshift distribution is not sculpted.
The input bias of SN~Ia host galaxies is set as $b=1.2$.  An independent measurement of $\Omega_{M_0}=0.3\pm0.005$
is included and is a non-trivial contributor to the $\gamma$ constraint.  Number densities are taken to be
direction-independent, neglecting the slight declination-dependence of SN-survey time windows
